---
title: "Describe data"
output: html_document
date: "2025-03-10"
---

## Introduction
This document explores the diabetes dataset by loading it, examining its structure, and identifying unique values for key variables.

## Load Data
```{r}
diabetesData <- read.csv("C:/Users/shanshan/Desktop/MH3511/Project/diabetes_012_health_indicators_BRFSS2015.csv", header=TRUE)
```

## Data Structure
```{r}
str(diabetesData)
```

## Unique Values in Key Variables

```{r}
# Diabetes status
diabetes_012 <- diabetesData[,1]
unique(diabetes_012)

# High blood pressure
high_bp <- diabetesData[,2]
unique(high_bp)

# High cholesterol
high_chol <- diabetesData[,3]
unique(high_chol)

# Cholesterol check
chol_check <- diabetesData[,4]
unique(chol_check)

# Body Mass Index (BMI)
bmi <- diabetesData[,5]
unique(bmi)

# Smoker status
smoker <- diabetesData[,6]
unique(smoker)

# Stroke occurrence
stroke <- diabetesData[,7]
unique(stroke)

# Heart disease or attack
heart_disease_or_attack <- diabetesData[,8]  # Fixed incorrect index
unique(heart_disease_or_attack)

# Physical activity
phys_activity <- diabetesData[,9]
unique(phys_activity)

# Fruit consumption
fruits <- diabetesData[,10]
unique(fruits)

# Vegetable consumption
veg <- diabetesData[,11]
unique(veg)

# Heavy alcohol consumption
hvy_alcohol_consump <- diabetesData[,12]
unique(hvy_alcohol_consump)

# Access to healthcare
any_healthcare <- diabetesData[,13]
unique(any_healthcare)

# No doctor due to cost
no_doc_bc_cost <- diabetesData[,14]
unique(no_doc_bc_cost)

# General health
gen_hlth <- diabetesData[,15]
unique(gen_hlth)

# Mental health
ment_hlth <- diabetesData[,16]
unique(ment_hlth)

# Physical health
phys_hlth <- diabetesData[,17]
unique(phys_hlth)

# Difficulty walking
diff_walk <- diabetesData[,18]
unique(diff_walk)

# Sex
sex <- diabetesData[,19]
unique(sex)

# Age
age <- diabetesData[,20]
unique(age)

# Education level
edu <- diabetesData[,21]
unique(edu)

# Income level
inc <- diabetesData[,22]
unique(inc)
```

## Conclusion
This analysis provides an initial overview of the dataset, ensuring that all key variables are correctly loaded and their unique values identified.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
